---
title: "Regression Models Course Project"
author: "Abhishek Kumar"
date: "22 September 2020"
output:
  word_document: default
  html_document: default
---

## Data Processing

In `mtcars` dataset, the variable `mpg` represents the Miles per (US) gallon and 
the `am` variable represents the tranmission where 0 represent the *automatic* and 
1 represent the *manual* transmission. 

```{r results='hide'}
library(datasets); library(ggplot2)
data("mtcars"); dat <- mtcars; str(mtcars)
dat$cyl <- as.factor(dat$cyl); dat$vs <- as.factor(dat$vs)
dat$am <- as.factor(dat$am); dat$gear <- as.factor(dat$gear)
dat$carb <- as.factor(dat$carb)
```

## Exploratory Analysis

```{r}
t.test(mpg~am, dat)$p.value
```

The t-test suggests that data provide sufficient evidence in favour of alternative
hypothesis i.e. there are significant differences as p-value (0.001) is less than 0.05.


## Regression Analysis

* **Model Selection:** The `step()` function in R uses the AIC for selection of best models 
i.e. with higher AIC values.Here, I will use a backward model 
selection strategy because it will prevent loss of any variable as we will start 
from a model which include all the predictors.

```{r}
mdl.full <- lm(mpg~., data = dat)
mdl.step <- step(mdl.full, direction = "backward", trace = FALSE)
summary(mdl.step)$coef
```

* **Coeffiecient Interpretation:** *This `mdl.step` model suggests that If all other 
variables are held constant, the `mpg` is expected to be 1.809 times higher for 
manual tranmission (`am1`) than the automatic transmission (`am0`) on average.*

* **Quantification of Uncertainity:**  

```{r}
confint(mdl.step, parm = "am1", level = 0.95)
```

Thus, our step model (`mdl.step`) suuggests that If all other variables are held 
constant,the `mpg` is expected to be 1.061 times lower to 4.679 times higher for 
manual tranmission (`am1`) than the automatic transmission (`am0`) on average at
95% significance level. since this is a relatively broader interval, the uncertainity 
in prediction can be moderate to high.

* **Model Diagnostics:** 

```{r fig.height=4, fig.width=8}
par(mfrow = c(1, 2)); plot(mdl.step, which = 1:2)
```

The above plots illustrates the model assumptions. The residuals vs Fitted plot
suggests moderate linear association as 3 points can be considered as ouliers. The
assumption of normaility is also fairly met since some points at the upper end are 
way from the normal line.

## Executive Summary

The miles per gallon (mpg) is an important parameter of fuel economy. Our analysis 
suggests that mpg is associated with number of cylinder (cyl), engine capacity 
horse power (`hp`), weight (`wt`) and transmission type (`am`). Although the 
transmission type is not a significant predictor of `mpg`, it has influence on
`mpg`. Statistical evidence supports that manual tranmission is more likely to have 
higher `mpg` on average, however, there is moderate to large uncertainity in the
coefficients suggesting that predictions may not be accurate.
