---
title: "Statistical Inference Project Part 2"
author: "Abhishek Kumar"
date: "9 August 2020"
output:
  word_document: default
  html_document: default
---

# Part 2: Basic Inferential Data Analysis

## Overview

Now in the second portion of the project, we're going to analyze the ToothGrowth 
data in the R datasets package. This analysis includes some basic exploratory analysis
specifically summary and hypothesis testing confidence intervals.

```{r}
library(dplyr)
library(ggplot2)
library(scales)
```


**1. Load the ToothGrowth data and perform some basic exploratory data analyses**

```{r}
data("ToothGrowth")
str(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
```

**About the data**

This data comes from a study which analyzed the effect of supplement (Vitamin C) 
on length of tooth in Guinea Pigs. This data has 60 observations and 3 variables 
(len, supp and dose). Each of 10 guinea 
pigs were given three Vitamin C dosage levels (0.5, 1, and 2 mg) with two delivery 
methods (orange juice or ascorbic acid). So, the data contains 60 observations 
of 3 variables.

* len : Tooth length
* supp : Supplement type (VC or OJ)
* dose : Dose in milligrams

## Exploratory Data Analysis

**2. Provide a basic summary of the data.**

```{r}
supp_summ <- ToothGrowth %>% group_by(Factor = supp) %>% summarise(mean = mean(len), 
                                                   sd = sd(len),
                                                   median = median(len),
                                                   minimum = min(len),
                                                   maximum = max(len),
                                                   IQR = IQR(len))

dose_summ <- ToothGrowth %>% group_by(Factor = as.factor(dose)) %>% summarise(mean = mean(len), 
                                                   sd = sd(len),
                                                   median = median(len),
                                                   minimum = min(len),
                                                   maximum = max(len),
                                                   IQR = IQR(len))

rbind(supp_summ, dose_summ)
```

The basic summary indicates that the tooth length was ranged from 4.2 to 33.9 with a 
mean of 18.81 and Median of 19.25. Among groups the data is shown in above table.


**3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.**

i. Which supplementary method is more effective for tooth growth?

$H_0: \mu_{OJ} = \mu_{VJ}; \quad H_A: \mu_{OJ} \ne \mu_{VJ}$

```{r}
OJ <- (ToothGrowth %>% filter(supp == "OJ"))$len
VC <- (ToothGrowth %>% filter(supp == "VC"))$len

t.test(OJ, VC, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```

```{r}
ggplot(ToothGrowth, aes(x = supp, y = len, fill = supp)) + 
      geom_boxplot(show.legend = FALSE) + 
      scale_fill_brewer(palette = "Set3") +
      theme_bw() + ylim(0,40) +
      labs(x = "Supplement Type", y = "Tooth Length (mm)",
           title = "Effect of Supplement type on Tooth Growth") +
      geom_label(x = 1, y = 37.5, fill = "gray95", size = 5,
                label = "paste(italic(t), \"= 1.915\", \", \" , italic(p), \"= 0.06\")", 
                parse = TRUE) + 
      geom_label(x = 2, y = 37.5, fill = "gray95", size = 5,
                label = "paste(\"95% CI\", \"= -0.171 to 7.571\")", 
                parse = TRUE)
```

So, the 95% confidence interval include zero suggesting that the null hypothesis
($H_0$) is true at this confidence interval and therefore, the alternative 
hypothesis ($H_A$) is rejected.

ii. Which dose is more effective for tooth growth?

Lets consider that the doses 0.5, 1, and 2 represents the levels as Low, Medium and
high. 

Hypotheses:

$H_0: \mu_{Low} = \mu_{Medium} = \mu_{High}; \quad \\ 
H_A: \mu_{Low} \ne \mu_{Medium} \ne \mu_{High}$

```{r}
Low <- (ToothGrowth %>% filter(dose == "0.5"))$len
Medium <- (ToothGrowth %>% filter(dose == "1"))$len
High <- (ToothGrowth %>% filter(dose == "2"))$len

t.test(Low, Medium, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(Low, High, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
t.test(High, Medium, paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```

Data | t | df | p-value | CI | x_mean | y_mean
-----|---|----|---------|----|--------|-------
Low vs Medium | -6.477 | 37.986 | <0.000 | -11.98 to -6.276| 10.60 | 19.73
Low vs High | -11.799 | 36.883 | <0.000 | -18.16 to -12.834| 10.60 | 26.10
High vs Medium | 4.901 | 37.101 | <0.000 | 3.734 to 8.996| 26.100 | 19.73

So, the 95% confidence interval did not included zero suggesting that the null hypothesis
($H_0$) is not true at this confidence interval and therefore, the alternative 
hypothesis ($H_A$) is accepted. Also, all the comparisons indicate a highly significant 
p-value showing the significant differences in the mean of the samples.

```{r}
ggplot(ToothGrowth, aes(x = as.factor(dose), y = len, fill = as.factor(dose))) + 
      geom_boxplot(show.legend = FALSE, notch = TRUE, size = 0.75) + 
      scale_fill_brewer(palette = "Set3") +
      theme_bw() + ylim(0,40) +
      labs(x = "Dose (mg)", y = "Tooth Length (mm)",
           title = "Effect of Dose on Tooth Growth")
```

### 4. State your conclusions and the assumptions needed for your conclusions.

#### Conclusions

The present analysis conclude that the tooth growth is not affected by supplement 
type (OJ or VC) at the 5% confidence intervals. However, the doses has the significant
effects on the tooth growth, as the mean of doses was significantly differed among the 
dose concentrations.

#### Assumptions

The analysis used the t-test for hypothesis testing. So it was assumed that:

1. The population from which samples are drawn is normally distributed.

2. The tests also assumed that variances are equally distributed or homogenous.

3. The samples were randomly drawn and represents the population.

4. The samples were independent of each other.